---
title: "Do you have a Date?"
author: "Learn how to work with dates using the lubridate package"
date: The following slides contain **new** commands that are not included in the RStudio
  codes and functions handout.
output:
  slidy_presentation: default
  pdf_document: default
---

```{r setup, include=FALSE}
```

```{r, eval=TRUE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(mobilizr)
library(lubridate)
```

## Installing packages

-   Click on the `Packages` tab and check if the lubridate package is installed.
    -   If it is not installed click on `Install`:

        -   type lubridate in the Packages box

        -   click Install

    -   Check off the box next to lubridate to load the package.
-   You are now ready to use lubridate!

## Lubridate basics

-   Run the code below using today's date in the form `month` `day` `year` as the argument

    ```{r, eval=FALSE}
    class("____________")

    ```

-   Run the code below using today's date as the argument again

    ```{r, eval=FALSE}
    class(mdy("_____________"))
    ```

<!-- -->

-   What was the output? What is the function mdy() doing?

-   What information will the function `wday()` extract from the date? Check and see if you were correct.

    ```{r, eval=FALSE}
    wday(mdy("_____________"))
    ```

-   Include the argument `label = TRUE` in the `wday()` function

The `mdy()` and `wday()` functions are functions in the `lubridate` package that help us extract information from dates. Let's get some more practice.

## The babies data

-   Use the `data()` command to import the *Births78* data.

-   The `wday`, `year`, `month`, `day_of_year`, `day_of_month` and `day_of_week` variables were all extract from the `date` column.

-   You are going to do an exercise where you recreate the *Births78* data set given only the `date` and `births` variables.

-   Run the code below to make a copy of the *Births78* data and simultaneously delete columns three through eight.

    ```{r, eval=FALSE}
    births_78<-select(Births78, -c(3:8))

    ```

-   Use the `wday()`function along with the `mutate()` function to create the `day_of_week` variable. Check your work against the original `Births78` dataset.

    ```{r, eval=FALSE, echo=TRUE}
    births_78<- _________(births_78, ___________ = _________(date))

    ```

    ```{r, eval=FALSE, echo=FALSE}
    births_78<-mutate(births_78, day_of_week=wday(date))

    ```

-   Run the code below inserting the variable name from the original *Births78* data that will be generated.

    ```{r, eval=FALSE}
    births_78<-mutate(births_78, _________ = wday(date, label= TRUE ))
    ```

## Lubridate commands

-   Go to the *Files* tab and open the *lubridate cheat sheet*

-   Refer to the functions in the GET AND SET COMPONENTS to determine which commands will extract the `year`, `month`, `day_of_year`, and `day_of_month` from the date variable.

    ```{r, eval= FALSE, echo = FALSE}
    births_78<-mutate(births_78, year = year(date))
    ```

    ```{r, eval= FALSE, echo = FALSE}
    births_78<-mutate(births_78, month = month(date))
    ```

    ```{r, eval= FALSE, echo = FALSE}
    births_78<-mutate(births_78, day_of_year = yday(date))
    ```

    ```{r, eval= FALSE, echo = FALSE}
    births_78<-mutate(births_78, day_of_month = day(date))
    ```

## On your own

-   Use either the `flights` dataset or the `crime` dataset and extract a useful variable from the date variable to perform analysis on.
